On the combination of Set Functions and Failures

### 2016-06-06 docs/src/extensions.tex
Manual updated, documentation of tools moved into currytools repo
Updated description of functional patterns w.r.t. syntactical restrictions

Functional patterns \cite{AntoyHanus05LOPSTR} are a useful extension
to code operations in a more readable way. Furthermore,
defining operations with functional patterns avoids problems
caused by strict equality (\ccode{=:=}) and leads to programs
that are potentially more efficient.

@@

In general, a pattern of the form \code{($f$ $t_1$\ldots$t_n$)} ($n>0$)
is interpreted as a functional pattern if $f$ is not a visible constructor
but a defined function that is visible in the scope of the pattern.

@@

\input{fieldlabels}

>>>>

Functional patterns \cite{AntoyHanus05LOPSTR} are a useful extension
to implement operations in a more readable way. Furthermore,
defining operations with functional patterns avoids problems
caused by strict equality (\ccode{=:=}) and leads to programs
that are potentially more efficient.

@@

In general, a pattern of the form \code{($f$ $t_1$\ldots$t_n$)} for $n>0$
(or of the qualified form \code{($M.f$ $t_1$\ldots$t_n$)} for $n \geq 0$)
is interpreted as a functional pattern if $f$ is not a visible constructor
but a defined function that is visible in the scope of the pattern.
Furthermore, for a functional pattern to be well-defined,
there are two additional requirements to be satisfied:

\begin{enumerate}

\item
If a function $f$ is defined by means of a functional pattern $fp$,
then the evaluation of $fp$ must not depend on $f$, i.e.,
the semantics of a function defined using functional patterns
must not (transitively) depend on its own definition.
This excludes definitions such as
\begin{curry}
(xs ++ ys) ++ zs = xs ++ (ys ++ zs)
\end{curry}
and is necessary to assign a semantics to funtions employing functional patterns
(see \cite{AntoyHanus05LOPSTR} for more details).

\item
Only functions that are globally defined may occur inside a functional pattern.
This restriction ensures that no local variable
might occur in the value of a functional pattern,
which might lead to a wrong semantics.
Consider, for instance, the following (complicated) equality operation
\begin{curry}
eq :: a -> a -> Bool
eq x y = h y
  where
  g True  = x
  h (g a) = a
\end{curry}
where the locally defined function \code{g} occurs in the functional
pattern \code{g a} of \code{h}.
Since \code{g a} evaluates to the value of \code{x} whereas \code{a}
is narrowed to \code{True}, the call \code{h y} now evaluates to \code{True}
if the value of \code{y} equals the value of \code{x}.
This requires a strict unification between \code{x} and \code{y},
so that a equivalent definition without functional patterns would be:
\begin{curry}
eq :: a -> a -> Bool
eq x y = h y
  where
  h x1 | x =:= x1 = True
\end{curry}
However, this implies that variables occuring in the value
of a functional pattern imply a strict unification
if they are defined in an outer scope,
whereas variables defined \emph{inside} a functional pattern
behave like pattern variables.
In consequence, the occurrence of variables from an outer scope
inside a functional pattern might lead to an unintuitive behavior,
which is why this is excluded by a restriction to globally defined functions.
Note that this does not exclude a functional pattern inside a local function,
which is still perfectly reasonable.
\end{enumerate}

@@
